---
title: "Homework-2"
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo=TRUE)
```


## Library calls
```{r}
library(tidyverse)
library(readxl)
```


## Problem 1

```{r}
# read csv file
nyc_transit_df = read_csv("NYC_Transit_Subway_Entrance_And_Exit_Data.csv")
#clean names
nyc_transit_df = janitor::clean_names(nyc_transit_df)
```
```{r}
# retain certain columns in df
nyc_transit_df = nyc_transit_df %>% select(line, station_name, station_latitude, station_longitude, route1, route2, route3, route4, route5, route6, route7, route8, route9, route10, route11, entry, entrance_type, vending, ada) 
```

```{r}
# recode entry variable from character to logical type
nyc_transit_df$entry = ifelse(nyc_transit_df$entry == "YES", TRUE, FALSE)
```

The nyc_transit_df datset contains variables that provide information about the NYC transit system. Some key variables include the subway line, the station name and the routes that the line serves. In order to clean this dataset, I cleaned the variable names, subsetted some of the columns and converted the entry variable to logical type. This dataset is not tidy because the route name and route line variables can be further concatenated using the pivot_longer function. 


```{r}
#Q: How many distinct stations are there? 
nyc_transit_df %>% 
distinct(line, station_name) %>% 
count()
#A: There are 465 distinct stations in NYC. 
```

```{r}
#Q: How many stations are ADA compliant?
nyc_transit_df %>% 
filter(ada == "TRUE") %>% 
distinct(line, station_name) %>%
count()
#A: 84 stations are ADA compliant. 
```
```{r}
#Q: What proportion of station entrances / exits without vending allow entrance?
prop1 <- nyc_transit_df %>% 
filter(vending == "YES") %>%
count(entry) %>%
mutate(prop = n/sum(n))
#A: 99.94% of station entrances / exits without vending vending allow entrance.
```

```{r}
#Reformat data so that route number and route name are distinct variables.
nyc_transit_df$route8 <- as.character(nyc_transit_df$route8)
nyc_transit_df$route9 <- as.character(nyc_transit_df$route9)
nyc_transit_df$route10 <- as.character(nyc_transit_df$route10)
nyc_transit_df$route11 <- as.character(nyc_transit_df$route11)

nyc_transit_df = 
  pivot_longer(
    nyc_transit_df, 
    route1:route11,
    names_to = "route_number",
    values_to = "route_name",
    names_prefix = "route"
  )
#Q:How many distinct stations serve the A train? 
nyc_transit_df %>%
filter(route_name == "A") %>%
distinct(line, station_name) %>%
count()
#A: 60 distinct stations serve the A train. 
```
```{r}
#Q: Of the stations that serve the A train, how many are ADA compliant?
nyc_transit_df %>%
filter(route_name == "A") %>%
filter(ada == "TRUE") %>%
distinct(line, station_name) %>%
count()
#A: Of the 60 stations that serve the A train, 17 are ADA compliant. 
```

## Problem 2

```{r}
# Read data for Mr. Trash Wheel
mr_trash_df = read_excel("trash_wheel_data.xlsx", sheet = 1, range = cell_cols("A:N"))
# clean names
mr_trash_df = janitor::clean_names(mr_trash_df)
#remove any rows without dumpster-specific data
mr_trash_df = mr_trash_df %>%
  filter(!is.na(dumpster))
#round number of sports balls to the nearest integer and converts the result to an integer variable
mr_trash_df$sports_balls <- round(mr_trash_df$sports_balls)
mr_trash_df$sports_balls <- as.integer(mr_trash_df$sports_balls)
#add new variable to indicate sheet number
mr_trash_df$sheet <- "Mr. Trash Wheel"
```

```{r}
# Read data for Professor Trash Wheel
prof_trash_df = read_excel("trash_wheel_data.xlsx", sheet = 2, range = cell_cols("A:M"))
# clean names
prof_trash_df = janitor::clean_names(prof_trash_df)
#remove any rows without dumpster-specific data
prof_trash_df = prof_trash_df %>%
  filter(!is.na(dumpster))

#add new variable to indicate sheet number
prof_trash_df$sheet <- "Professor Trash Wheel"

#combine datasets
prof_trash_df$year <- as.character(prof_trash_df$year)
joint_trash_df <- bind_rows(mr_trash_df, prof_trash_df)

```


```{r}
#Q: What was the total weight of trash collected by Professor Trash Wheel?
prof_trash_wt = joint_trash_df %>% 
  filter(sheet == "Professor Trash Wheel") %>%
  summarise(total_weight = sum(weight_tons))
#A: The total weight of trash collected by Professor Trash is 190.12 tons. 
```
```{r}
#Q: What was the total number of sports balls collected by Mr. Trash Wheel in 2020?
mr_sports_balls = joint_trash_df %>% 
  filter(sheet == "Mr. Trash Wheel" & year == 2020) %>%
  summarise(total_balls = sum(sports_balls))
#A: Mr. Trash Wheel collected a total of 856 sports balls in 2020. 
```
This dataset provides information about trash collected at various dumpster sites by Mr. Trash Wheel and Professor Trash Wheel. The dataset has `r ncol(joint_trash_df)` variables and `r nrow(joint_trash_df)` observations. It tells us about the weight of the trash collected on specific days and also provides information about the number of plastic bottles, glass bottles, cigarette butts, and other constituents of the trash. The total weight of trash collected by Professor Trash Wheel is `r prof_trash_wt` tons and Mr. Trash Wheel collected a total of `r mr_sports_balls` sports balls in 2020. 

## Problem 3
```{r}
# read and clean the data in pols-month.csv
pols_month_df = read_csv("problem3_data/pols-month.csv")
#clean names
pols_month_df = janitor::clean_names(pols_month_df)
# Use separate() to break up the variable mon into integer variables year, month, and day AND replace month number with month name
pols_month_df = pols_month_df %>% 
  separate(mon, into = c("year", "month", "day"), sep = "-") %>%
  arrange(year, month) %>%
  mutate(month = recode(month, "01" = "January", "02" = "February", 
                        "03" = "March", "04" = "April", "05" = "May", 
                        "06" = "June", "07" = "July", "08" = "August",
                        "09" = "September", "10" = "October", 
                        "11" = "November", "12" = "December"))

#create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable
pols_month_df$president = paste(pols_month_df$prez_dem, ",", pols_month_df$prez_gop)

pols_month_df = subset(pols_month_df, select = -c(prez_dem, prez_gop, day))
#arrange df according to year and month
```

```{r}
# read and clean the data in snp.csv
snp_df = read_csv("problem3_data/snp.csv")
#clean names
snp_df = janitor::clean_names(snp_df)
# Use separate() to break up the variable mon into integer variables year, month, and day
snp_df = snp_df %>% 
  separate(date, into = c("month", "day", "year"), sep = "/") %>%
  arrange(year, month) %>%
  select(-day) 

#organize so that year and month are the leading columns
snp_df <- snp_df[, c(2,1,3)]
#edit year column
snp_df$year[1:187] = paste0("20", snp_df$year[1:187])
snp_df$year[188:787] = paste0("19", snp_df$year[188:787])
#make month numeric type
snp_df$month <- as.numeric(snp_df$month)
#arrange snp_df again according to year and replace month number with month name
snp_df = snp_df %>%
  arrange(year, month) %>%
  mutate(month = recode(month, "1" = "January", "2" = "February", 
                        "3" = "March", "4" = "April", "5" = "May", 
                        "6" = "June", "7" = "July", "8" = "August",
                        "9" = "September", "10" = "October", 
                        "11" = "November", "12" = "December"))
```

```{r}
# read and clean the data in unemployment.csv
unemployment_df = read_csv("problem3_data/unemployment.csv")
#clean names
unemployment_df = janitor::clean_names(unemployment_df)
#pivot_longer month
unemployment_df = 
  pivot_longer(
    unemployment_df, 
    jan:dec,
    names_to = "month",
    values_to = "unemployment_value"
  )
unemployment_df = unemployment_df %>%
  mutate(month = recode(month, "jan" = "January", "feb" = "February", 
                        "mar" = "March", "apr" = "April", "may" = "May", 
                        "jun" = "June", "jul" = "July", "aug" = "August",
                        "sep" = "September", "oct" = "October", 
                        "nov" = "November", "dec" = "December"))
```
```{r}
#combine datasets
pols_snp_joint = left_join(pols_month_df, snp_df)
unemployment_df$year = as.character(unemployment_df$year)
pols_snp_unemp_joint = left_join(pols_snp_joint, unemployment_df)
```
#Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).

The pols_snp_unemp_joint dataset can be used to see if the U.S. economy is affected by whether Republicans or Democrats are in office. This dataset was created by joining three datsets: (1) pols_month_df, which contains information about the number of Democrats and Republicans in office, (2) snp_df, which contains information about the closing price of the stock market and, (3) unemployment_df, which contains information about the unemployment rates. The resulting dataset has `r ncol(pols_snp_unemp_joint)` columns and `r nrow(pols_snp_unemp_joint)` rows. Its observations range from `r pols_snp_unemp_joint$year[1]` to `r pols_snp_unemp_joint$year[822]`. It has important variables such as the number of Democrats/ Republicans in the senate, the closing price of the stock market and the unemployment rate. 
